When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-22 doc/paper/vpd_et_paper_PG_v1.tex
half way done with morn writing, takinga  break and will come back

Figure \ref{term3}b shows the location of VPD$_{crit}$, as a function of $\sigma$ (uncertainty) and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. This motivates a thorough examination of how inclusion of uncertainty ($\sigma$) might change the form of the sign term, and in particular the location of VPD$_{crit)$ for each of our plant types. In Secion \ref{testing} we examine if and how uncertainty changes the analysis presented in this section, for each PFT.

@@

For DBF our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included.  The plots for DBF look as one would expect if one just added noise on top of our theory.

For CSH our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included.  The plots for CSH look as one would expect if one just added noise on top of our theory.

The theory does not match as well for CRO . Perhaps this should be expected. Our theory, for practical reasons, does not include important sources of variability (particularly between sites) for CRO. We do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis, or differences in irrigation practices between the sites. Given that the our theory does not do as well for CRO, we will refrain from firm statements about how CRO respond to changes in VPD.

The theory does not match as well for GRA. Perhaps this should be expected. Our theory, for practical reasons, does not include important sources of variability (particularly between sites) for GRA. We do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis, or differences in irrigation practices between the sites. Given that the our theory does not do as well for GRA, we will refrain from firm statements about how  GRA respond to changes in VPD.

ENF is a special case. The VPD dependence of the theory is not well captured by observations. However, the theory closely captures the intercept where $\frac{\partial \; ET}{\partial \; VPD} = 0$. Resolving this feature is the key to answering our question, so perhaps the theory is still useful for ENF for the purpose of addressing when VPD drives or reduces ET. It is also noteworthy that we used published constants for $g_1$, which determines the VPD dependence, from \cite{Lin_2015}. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori physical constants whenever possible.

While the above discussion shows that our theory has some limitations when applied to some sites, especially for crops and grasslands, another question could be, how does our theory's performance compare to often-used PET? Again, as in Section \ref{sign_func}, the PET response to changes in VPD will be constant at about 3.5. In Figure \ref{agu_fig} the ET reponse to shifts in dryness is always far below (more negative) 3.5 for observations, including uncertainty. So the PET reponse to atmospheric dryness is outside the range of all observable responses. This casts doubt on PET's usefulness for capturing the leading order behavior of water budgeting at the land surface. However our theory (Equation \ref{et}) could be used with published $uWUE$ and $g1$ as an alternative to PET to estimating land response and moisture budgeting.

>>>>

Figure \ref{term3}b shows the location of VPD$_{crit}$, as a function of $\sigma$ (uncertainty) and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. This motivates a thorough examination of how inclusion of uncertainty ($\sigma$) might change the form of the sign term, and in particular the location of VPD$_{crit}$ for each of our plant types. In Secion \ref{testing} we examine if and how uncertainty changes the analysis presented in this section, for each PFT.

@@

For DBF our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included, and the DBG figure looks as one would expect if one just added noise on top of our theory. The VPD dependence, given by the slope parameter ($g_1$), floows the median values of each box plot. Perhaps most importantly, the x-intercept, and thus VPD$_{crit}$, matches nearly exactly between the theory and the observations. Therefor the sign of the ET response to increases in VPD should be well matched, subject to the unavoidable constraints of noise, much of which may be from the observations themselves. The uncertainty is non negligable; there are bany observations in each box for which the the sign of observation is opposite the response predicted by the theory, but to leading order our theory matches the observations.

While CSH has a much different functional form of the sign term than DBF, CSH observations also match our theory to leading order. Again, the VPD dependence mostly determined from the slope parameter ($g_1$) closely matches the medians in the observation bins. The VPD-independent, strongly negative response, which is determined by $uWUE$, is also captured. For CSH, there is extremely rare occurence of positive ET response with VPD, even with uncertainty, so the sign of the theory almost always matches the sign of the observations, even with all of the uncertainty.

The theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similiar to the leading order behavior of the observations, but the entire curve is shifted towards postiive values, due to some error in $uWUE$. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependnce, is behavior that is impossible fo the theory to capture. This divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of variability (particularly between sites) for CRO. We do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis, or differences in irrigation practices between the sites and/or years. This deficiencies would explain the inability of our theory to match obsercations. For example, a superpositions of sites with C3 plants (low VPD) and C4 plants (high VPD) would explain observed shift back towards negative ET reponse at high VPD when all sites are binned together, as in Figure \ref{agu_fig}.

The theory for GRA suffers from similiar limitations as for CRO. The differences between the theory and the observations are more pronounced for GRA, but are explainable using similiar logic as that used for CRO. GRA will also have differences between sites in C3 vs C4 photosythesis, and irrigation practices, as well as strong seasonal variability in crop height and LAI. For GRA, the observations show a very different VPD dependence of response than the theory, with a strong tendency back toewards a negative ET reponse at high VPD which our theory cannot account for. As with CRO, this could be explained by a superposition of sites with C3 vs C4 sites. Unfortunately, because detailed site-level data of C3 vs C4 plant type, as well as irrigation practes, was difficult to obtain, we could not bin GRO and GRA further by irrigation and photosyethesis stragtegy. However, we hypothesize that the theory would much more closel match obsercations with photosythesis and irrigation are accounted for.

The comparison between the theory and obsercations for ENF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope paramter, is not well captured by observations. However, the theory closely captures the x-intercept and VPD$_{crit}$. So, the theory still captures when the ET reponse will be positive or negative, but is less accurate with the magnitude of the reponse. Therefor the theory is still useful for addressing when VPD drives or reduces ET. Also, it's with noting that we used published constants for $g_1$ \citep[from ][]{Lin_2015}, which determines the VPD dependence. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori physical constants whenever possible.

While the above discussion shows that our theory has some limitations when applied to some sites, especially for crops and grasslands, another question could be, how would using PET, as many hydrometeorological studies and applications do, compare with our theory.  Again, as in Section \ref{sign_func}, the PET response to changes in VPD will be constant at about 3.5. In Figure \ref{agu_fig} the ET reponse to shifts in dryness is always far more negative than 3.5 for the observations, even with all uncertainty included. So the PET reponse to atmospheric dryness is outside the range of all observable responses. This casts doubt on PET's usefulness for capturing the leading order behavior of water budgeting at the land surface. However our theory (Equation \ref{et}) could be used with published $uWUE$ and $g1$ as an alternative to PET to estimating land response and moisture budgeting.
